Combinatorial interpretations of holonomic recurrences

### 2021-09-04 article/main.tex
Fixups

\section{Overview}%

Every recursive structure can be described using a generating function, which can be then transformed into holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations, like adding some vertices to the smaller trees. Although, there is no known method of interpreting it in the language of the combinatorial structures.

Initial inspiration for the thesis came from publication \textit{A natural counting of lambda terms}\cite{zigzag}, which shows bijection between seemingly different tree structures, which are lambda trees, black-white trees and zigzag free trees. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structure itself.

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an approach that helped us to find an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation.

In the section\ \ref{sec:introduction} we establish notation nad introduce concepts that will be using throughout the thesis.

Next, in the section \ref{sec:known_results}, we present known results. They demonstrate combinatorial interpretations that reflect different approaches.

In the section\ \ref{sec:approaches} we show approaches to the problem that helped people achieve their results. All of them were tried in order to find the answer to the original problem of interpreting holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately none of them helped us enough to get the final result. They are proven to be helpful when we are stuck though as most of them are able to divide problems into smaller parts and detect corner cases.

Section\ \ref{sec:brutal_iterative_approach} presents our final approach to the original problem, i.e.\ computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is able to provide us with the full solution, but we were not able to achieve that due to the limited time that was available to us.

Finally, in section\ \ref{sec:further_work}, we present our view on what can be done with the problem in the future. We believe that one day the problem will be fully solved and a more generic approach to interpreting holonomic recurrences will emerge.

@@

@@

i.e.\ stands for Latin id est, which means \textit{that is}.
\todo{It should be obvious, but people mix them up}

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define a full binary tree recursively as a node having no children (i.e.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

\subsection{Generating functions}%
\label{sub:generating_functions}
\todo{Make the section simpler}

@@

\begin{definition}
The ordinary generating function of sequence \(A = (a_i)_{i=0}^{\infty}\) is defined as:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i\]
\end{definition}

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without the need of digging into details. The tool we will be using extensively in this work is the Maple package called GFUN\cite{gfun}. If you want to learn more about algorithms used in the package, we strongly encourage reading the referenced article.

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, i.e.\ none of them satisfies a linear recurrence equation with polynomial coefficients\cite{nonholo}.
\end{example}

@@

The equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def:holo_1}.\cite{complexity}

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

Combinatorial interpretation is just a fancy name for something you probably know and use. Eg.\ if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\) you interpret it as number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

@@

Let's start with probably the most known result, full binary trees\cite{binary}.

@@

Let's think, what would have changed if we introduce unary nodes to the binary case\cite{motzkin}. We will be counting edges. Note that a tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

And with a bit of Maple magic\cite{gfun}, we immediately get following recurrence:

@@

To present the interpretation in bijective language, we will have to introduce several types of the the nodes (you may want to revisit section \ref{sub:notation} Notation):

@@

The next case we will show is the case of Schröder trees\cite{schroder}.

@@

As the most common trees are binary ones, it may be worth an effort to find some structure based on binary trees that can represent the original structure (see section~\ref{sub:unary_binary_trees}).

@@

In section \ref{sub:lambda_terms} we will define trees of lambda terms. If we would assign following weights to nodes:

@@

Another method that may help finding the interpretation is inductive translation\cite{doron}. The method is based on the idea that most of the time there is an easy way to find some natural induction (not necessarily holonomic) that has trivial interpretation in bijective language.

@@

Using a little bit of Maple magic\cite{gfun}, we can easily achieve the following holonomic recurrence.

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate between nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, i.e.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, i.e.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, i.e.~is right child of its parent;
    \item node has its parent on the right, i.e.~is the left child of its parent.
\end{itemize}

@@

You may remember our assumptions in section \ref{sub:the_framework} about the discrimination cases. We allowed ourselves to apply visitors only to one of the left and right children, which is not handled by our holonomic recurrence as the number of such vertices does not depend only on \(n\).

@@

First thing to be done, that comes to mind immediately, is to finish the construction eliminating the cheat mentioned in section \ref{sub:the_cheat}.

@@

With the proper interpretation, it may be possible to create a sampling algorithm similar to Rémy's one for binary trees\cite{remy,note}. It would have to take under consideration that some trees are being removed by other classes, so the probabilities would have to be set carefully.

>>>>

\section{Overview}%
\todo{generating functions}
\todo{holonomic recurrences}
\todo{publications about Motzkin, etc.}
\todo{Lambda terms}
\todo{zigzag free}

Every recursive structure can be described using a generating function, which can be then transformed into a holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

The initial inspiration for the thesis came from the publication \textit{A natural counting of lambda terms}~\cite{zigzag}, which shows bijections among seemingly different tree structures, which are lambda trees, black-white trees\todo{cite} and zigzag free trees\todo{cite}. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structures itself.

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an approach that helped us to find an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation. We use the de Bruijn indices in order to get rid of letter-based variable names. Unary notation allows us to connect size of the tree to the values of the de Bruijn indices.

In Section~\ref{sec:introduction} we establish notation nad introduce concepts that will be used throughout the thesis.

Next, in Section~\ref{sec:known_results}, we present known results, ie. binary trees, unary-binary trees and Schröder trees. They demonstrate combinatorial interpretations that reflect different approaches.

In Section~\ref{sec:approaches} we show approaches to the problem that helped to achieve the aforementioned results. All of them were tried in order to find the answer to the original problem of interpreting holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately, none of them helped us enough to get the final result. Anyway, they are proven to be helpful when we are stuck, as most of them are able to divide problems into smaller parts and detect corner cases.

Section~\ref{sec:brutal_iterative_approach} presents our final approach to the original problem, i.e.\ computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is possible to extend this idea in order to obtain the full solution, but we were not able to achieve that due to the limited time that was available to us.

Finally, in Section~\ref{sec:further_work}, we present our view on what can be done with the problem in the future. We believe that one day the problem will be fully solved and a more generic approach to interpreting holonomic recurrences will emerge.

@@

i.e.\ stands for Latin id est, which means \textit{that is}.

@@

@@

We will be analyzing structures that can be expressed using recursive formulas. For example we can define a full binary tree recursively as a node having no children (i.e.\ being a leaf) or having exactly two subtrees being full binary trees. To express the disjoint alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

\subsection{Generating functions}%
\label{sub:generating_functions}
\todo{Source}
\todo{Make the section simpler}

@@

\begin{definition}
The ordinary generating function of sequence \(A = (a_i)_{i=0}^{\infty}\) is defined as a formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i .\]
\end{definition}

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without the need of digging into details. The tool we will be using extensively in this work is the Maple package called GFUN~\cite{gfun}. If you want to learn more about algorithms used in the package, we strongly encourage reading the referenced article.

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, i.e.\ none of them satisfies a linear recurrence equation with polynomial coefficients~\cite{nonholo}.
\end{example}

@@

The equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def:holo_1}~\cite{complexity}.

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

Combinatorial interpretation is just a fancy name for something you probably know and use. For example if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\) you interpret it as number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

@@

Let's start with probably the most known result, full binary trees~\cite{binary}.

@@

Let's think, what would have changed if we introduce unary nodes to the binary case~\cite{motzkin}. We will be counting edges. Note that a tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

And with a bit of Maple magic~\cite{gfun}, we immediately get following recurrence:

@@

To present the interpretation in bijective language, we will have to introduce several types of the the nodes (you may want to revisit Section~\ref{sub:notation} Notation):

@@

The next case we will show is the case of Schröder trees~\cite{schroder}.

@@

As the most common trees are binary ones, it may be worth an effort to find some structure based on binary trees that can represent the original structure (see Section~\ref{sub:unary_binary_trees}).

@@

In Section~\ref{sub:lambda_terms} we will define trees of lambda terms. If we would assign following weights to nodes:

@@

Another method that may help finding the interpretation is inductive translation~\cite{doron}. The method is based on the idea that most of the time there is an easy way to find some natural induction (not necessarily holonomic) that has trivial interpretation in bijective language.

@@

Using a little bit of Maple magic~\cite{gfun}, we can easily achieve the following holonomic recurrence.

@@

We also created a framework~\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate between nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, i.e.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, i.e.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, i.e.~is right child of its parent;
    \item node has its parent on the right, i.e.~is the left child of its parent.
\end{itemize}

@@

You may remember our assumptions in Section~\ref{sub:the_framework} about the discrimination cases. We allowed ourselves to apply visitors only to one of the left and right children, which is not handled by our holonomic recurrence as the number of such vertices does not depend only on \(n\).

@@

First thing to be done, that comes to mind immediately, is to finish the construction eliminating the cheat mentioned in Section~\ref{sub:the_cheat}.

@@

With the proper interpretation, it may be possible to create a sampling algorithm similar to Rémy's one for binary trees~\cite{remy,note}. It would have to take under consideration that some trees are being removed by other classes, so the probabilities would have to be set carefully.
